---
title: Proposed model for flu hospitalization
description: Notebook to compare hospitalization forecasting models.
output:
  html_document:
    code_folding: hide
---

### Summary

**Background:**

* Flu (and COVID) hospitalization collected centrally by the HHS starting
  this year (2021).
* The CDC is asking for forecasts of flu hospitalization incidence at the 
  weekly level.  Forecasts are due every Monday evening, for the following
  four ending Epiweeks.  This means that if Monday is the forecast date, 
  and we are predicting seven day averages, then we are to submit
  forecasts for `ahead = c(5, 12, 19, 26)`.
* Forecasts are to be submitted for the 50 US states, DC, VI, and PR; 
  as well as national-level forecasts comprised of those 53 state-level
  entities.  These distributional forecasts are to consist predictions at
  23 quantiles.


### Methods

**Inputs:**

* `hhs:confirmed_admissions_influenza_1d_prop_7dav`: this data describes
  population-normalized number of flu hospitalizations in each state
  ("flu hospitalization rate"), per day.
  It is normalized to the incidence ratio scale (count / population * 100,000).
* The data are smoothed by taking a **rolling seven-day average.**
* We include all **50 US states, DC, VI, and PR**.
* Here, we take forecast dates from **2021-10-01 to the latest available**.

**Outputs:**

* For each forecast date, we predict the flu hospitalization rate at 
  **aheads=5:28**.
* For each target date, at each of the **53 locations**, we produce a
  distributional forecast, predicting at the 23 COVIDhub quantiles.

**Model:**

Here we re-use the notation from the pnas-forecast paper.

Let $Y_{\ell, t}$ denote the 7-day trailing average of flu hospitalization
incidence rates in location $\ell$ and at time $t$.  Further,
let $X_{\ell, t}$ denote the 7-day trailing average of an auxiliary
indicator; here, the auxiliary indicator is outpatient doctor visits
with confirmed influenza, reported by Change Healthcare.  Separately for
each $a = 5, \dotsc, 28$, we learn a model of the form:
$$
\mathrm{Quantile}_\tau(Y_{\ell, t+a} | Y_{\ell, s}, s \leq t)
= \alpha^{a, \tau}
+ \sum_{j=0}^2 \beta_j^{a,\tau}Y_{\ell, t-7j}
+ \sum_{j=0}^J \gamma_j^{a,\tau}X_{\ell, t-7j},
$$
where we we take $J = -1, 0, 2$.

For a fixed target quantile $\tau$, this uses current flu hospitalization
rates, as well as seven- and fourteen-day lagged flu hospitalization
rates, in order to predict the $\tau$-quantile of case rates at
$a$ days ahead.

This model is fit using quantile regression separately for each
$\tau$ (with post hoc sorting and thresholding), using data from all
53 "states" and the most recent 21 days of training data.

```{r setup, message=FALSE, warning=FALSE}
library(covidcast)
library(dplyr)
library(tidyr)
library(tibble)
library(ggplot2)
library(Matrix)

source('common_params.R')
aheads=ahead;rm(ahead)
hhs_source = 'hhs'
flu_hosp_num  = 'confirmed_admissions_influenza_1d_7dav'
flu_hosp_prop = 'confirmed_admissions_influenza_1d_prop_7dav'
covid_hosp_num  = 'confirmed_admissions_covid_1d_7dav'
covid_hosp_prop = 'confirmed_admissions_covid_1d_prop_7dav'
jhu_source = 'jhu-csse'
covid_case_num  = 'confirmed_7dav_incidence_num'
covid_case_prop = 'confirmed_7dav_incidence_prop'
start_day =  '2020-05-01'
end_day =    last_eval_date
colors = 1:4
```
### Forecasting results

We trained two forecasters, using forecast dates from October 1, 2021 to
(latest possible).  For each forecast date, we predict 5 days ahead to 
28 days ahead.  Predictions are made for the 23 CDC quantiles.
Training was done on the rates scale; one can transform the predictions to
the counts scale (as requested by the CDC) post hoc.

Based on the following results, we recommend using the "AR3+CHNG3" model, 
which uses three lags of the response and the Change flu indicator.  Adding
the lagged indicator provides modest improvements in WIS, and noticeable 
improvements in coverage (although all models are severely undercovering).

```{r plot_results_wis, fig.width=12, fig.height=12, warning=FALSE, message=FALSE}

full_results = readRDS('results/results_23.RDS')

results = full_results %>% filter (
      !stringr::str_ends(forecaster, '_7'),
      !stringr::str_detect(forecaster, 'logtrans'),
    ) %>% mutate (
      forecaster=stringr::str_replace(forecaster, '_23', ''),
    )

state_pop = covidcast::state_census %>% tibble (
    ) %>% transmute (
      geo_value = stringr::str_to_lower(ABBR),
      pop = POPESTIMATE2019,
    ) %>% mutate (
      state_with_pop = sprintf('%s, %d', geo_value, pop),
      state_with_pop = factor(state_with_pop,
                      levels = state_with_pop[order(pop)])
    )


results = results %>% left_join (
      state_pop,
      by='geo_value',
    ) %>% mutate (
      pop_wis = wis * pop,
      pop_ae = ae * pop,
    )

results$forecaster = factor(results$forecaster,
                            levels=c('Baseline','AR3','AR3+CHNG1','AR3+CHNG3'))

mean_results = results %>% group_by (
        forecaster,
        ahead,
      ) %>% summarize (
        mean_wis = mean(wis),
        mean_pop_wis = mean(pop_wis),
        mean_cov80 = mean(cov80),
        median_wis = median(wis),
        median_pop_wis = median(pop_wis),
      ) %>% ungroup
relative_mean = inner_join (
      mean_results %>% filter (
        forecaster == 'Baseline'
      ) %>% select (
        -forecaster,
      ),
      mean_results,
      by=c('ahead'),
      suffix=c('_Baseline', '')
    ) %>% transmute (
      ahead,
      forecaster,
      relative_mean_wis = mean_wis/mean_wis_Baseline,
      relative_median_wis = median_wis/median_wis_Baseline,
    )

plt_mean_wis = mean_results %>% ggplot (
      aes(x=ahead,
          y=mean_wis,
          color=forecaster)
    ) + geom_point (
    ) + geom_line (
    ) + scale_color_manual(
      values=colors,
    ) + geom_vline (
      xintercept=c(5, 12, 19, 26),
      linetype='dotted',
    ) + xlab (
      'Mean WIS'
    ) + theme(
      legend.position="bottom"
    )

plt_mean_pop_wis = mean_results %>% ggplot (
      aes(x=ahead,
          y=mean_pop_wis,
          color=forecaster)
    ) + geom_point (
    ) + geom_line (
    ) + scale_color_manual(
      values=colors,
    ) + geom_vline (
      xintercept=c(5, 12, 19, 26),
      linetype='dotted',
    ) + xlab (
      'Mean WIS (population-weighted)'
    ) + theme(
      legend.position="bottom"
    )

plt_relative_mean = relative_mean %>% ggplot (
      aes(x=ahead,
          y=relative_mean_wis,
          color=forecaster)
    ) + geom_point (
    ) + geom_line (
    ) + scale_color_manual(
      values=colors,
    ) + geom_vline (
      xintercept=c(5, 12, 19, 26),
      linetype='dotted',
    ) + xlab (
      'Mean WIS (relative to mean WIS of Baseline)'
    ) + theme(
      legend.position="bottom"
    )

plt_cov80 = mean_results %>% ggplot (
      aes(x=ahead,
          y=mean_cov80,
          color=forecaster)
    ) + geom_point (
    ) + geom_line (
    ) + scale_color_manual(
      values=colors,
    ) + geom_vline (
      xintercept=c(5, 12, 19, 26),
      linetype='dotted',
    ) + geom_hline (
      yintercept=0.8,
      linetype='dotted',
    ) + xlab (
      'Prop. 80% intervals that covered truth'
    ) + theme(
      legend.position="bottom"
    )

gridExtra::grid.arrange(plt_mean_wis,
                        plt_mean_pop_wis,
                        plt_relative_mean,
                        plt_cov80,
                        nrow=2)

```


```{r plot_results_wis_diff, fig.width=12, fig.height=12, warning=FALSE, message=FALSE}

results_diff = results %>% filter (
      ahead %in% c(5, 12, 19, 26),
      forecaster %in% c('Baseline', 'AR3+CHNG3'),
    ) %>% group_by (
      geo_value,
      ahead,
      forecaster,
    ) %>% summarize (
      mean_wis = mean(wis),
      mean_cov80 = mean(cov80),
    ) %>% ungroup (
    ) %>% inner_join (
      state_pop,
      by='geo_value',
    )

plt = ggplot(results_diff,
             aes(x=pop,
                 y=mean_wis,
                 color=forecaster),
    ) + geom_point (
    ) + scale_color_manual (
      values = c(1, 4),
    ) + facet_wrap (
      vars(ahead),
      nrow=2
    ) + xlab (
      'Mean WIS'
    ) + scale_x_log10 (
    ) + theme (
      legend.position='bottom',
    )

plt

plt = ggplot(results_diff,
             aes(x=pop,
                 y=mean_cov80,
                 color=forecaster),
    ) + geom_point (
    ) + scale_color_manual (
      values = c(1, 4),
    ) + facet_wrap (
      vars(ahead),
      nrow=2
    ) + geom_hline (
      yintercept=0.8,
      linetype='dotted',
    ) + xlab (
      'Prop. 80% intervals that covered truth'
    ) + scale_x_log10 (
    ) + theme (
      legend.position='bottom',
    )

plt
```
