---
title: Proposed model for flu hospitalization
description: Notebook to compare hospitalization forecasting models.
output:
  html_document:
    code_folding: hide
---

### Summary

**Background:**

* Flu (and COVID) hospitalization collected centrally by the HHS starting
  in 2021.
* The CDC is asking for forecasts of flu hospitalization incidence at the 
  weekly level.  Forecasts are due every Monday evening, for the following
  four ending Epiweeks.  This means that if Monday is the forecast date, 
  and we are predicting seven day averages, then we are to submit
  forecasts for `ahead = c(5, 12, 19, 26)`.
* Forecasts are to be submitted for the 50 US states, DC, VI, and PR; 
  as well as national-level forecasts comprised of those 53 state-level
  entities.  These distributional forecasts are to consist predictions at
  23 quantiles.
* Here, we consider a [baseline ("flat-line") forecaster](https://cmu-delphi.github.io/covidcast/evalcastR/reference/baseline_forecaster.html)
  based on the [COVID Forecast Hub](https://github.com/reichlab/covid19-forecast-hub)
  baseline, an autoregressive model based solely on HHS data, and
  autoregressive models that augment the HHS data with
  auxiliary information from Change Healthcare.
* **We recommend that at "one week" ahead (ahead=5), we use the
  "flat-line" forecaster; but at two or more weeks ahead, we use
  an autoregressive model with Change Healthcare features.**


### Methods

**Inputs:**

* The principal autoregressive feature is
  `hhs:confirmed_admissions_influenza_1d_prop_7dav`.  This data describes
  population-normalized number of flu hospitalizations in each state
  ("flu hospitalization rate"), per day.
  It is normalized to the incidence ratio scale (count / population * 100,000)
  and smoothed by taking a rolling seven-day average.
  [Documentation in COVIDcast](https://cmu-delphi.github.io/delphi-epidata/api/covidcast-signals/hhs.html).
* We also include
  `chng:smoothed_adj_outpatient_flu`: this data uses Change Healthcare 
  claims to estimate the percentage of outpatient doctor visits with
  confirmed influenza (day-of-week effects removed).
  [Documentation in COVIDcast](https://cmu-delphi.github.io/delphi-epidata/api/covidcast-signals/chng.html).
* We include all **50 US states, DC, VI, and PR**.
* Here, we take forecast dates from **2021-10-01 to the latest
  available**.  This notebook uses data as of December 30, 2021.
* **Important note:** the earliest issue available for the Change Healthcare 
  data is December 6, 2021, so this is trained with "finalized" data 
  as-of the evaluation date on which the models were trained.  The HHS features
  are taken with "vintage" data as-of each forecasting date.  The use of
  "finalized" Change Healthcare data may lead to optimistic performance.
  We can switch to "vintage" data once backissues are restored to 
  COVIDcast (or after a substantial amount of time has passed, after 
  which we can ignore forecasting dates before December 6th).


**Outputs:**

* For each forecast date, we predict the flu hospitalization rate at 
  **aheads=5:28**.
* For each target date, at each of the **53 locations**, we produce a
  distributional forecast, predicting at the 23 COVIDhub quantiles.

**Model:**

Here we use the same notation as in the
[PNAS forecasing evaluation paper](https://www.pnas.org/content/118/51/e2111453118).

Let $Y_{\ell, t}$ denote the 7-day trailing average of flu hospitalization
incidence rates in location $\ell$ and at time $t$.  Further,
let $X_{\ell, t}$ denote the 7-day trailing average of an auxiliary
indicator; here, the auxiliary indicator is outpatient doctor visits
with confirmed influenza, reported by Change Healthcare.  Separately for
each $a = 5, \dotsc, 28$, we learn a model of the form:
$$
\mathrm{Quantile}_\tau(Y_{\ell, t+a} | Y_{\ell, s}, s \leq t)
= \alpha^{a, \tau}
+ \sum_{j=0}^2 \beta_j^{a,\tau}Y_{\ell, t-7j}
+ \sum_{j=0}^J \gamma_j^{a,\tau}X_{\ell, t-7j},
$$
where we we take $J = -1, 0, 2$.

For a fixed target quantile $\tau$, this uses current flu hospitalization
rates, as well as seven- and fourteen-day lagged flu hospitalization
rates, in order to predict the $\tau$-quantile of case rates at
$a$ days ahead.

This model is fit using quantile regression separately for each
$\tau$ (with post hoc sorting and thresholding), using data from all
53 "states" and the most recent 21 days of training data.

```{r setup, message=FALSE, warning=FALSE}
library(covidcast)
library(dplyr)
library(tidyr)
library(tibble)
library(ggplot2)
library(Matrix)

source('common_params.R')
aheads=ahead;rm(ahead)
hhs_source = 'hhs'
flu_hosp_num  = 'confirmed_admissions_influenza_1d_7dav'
flu_hosp_prop = 'confirmed_admissions_influenza_1d_prop_7dav'
covid_hosp_num  = 'confirmed_admissions_covid_1d_7dav'
covid_hosp_prop = 'confirmed_admissions_covid_1d_prop_7dav'
jhu_source = 'jhu-csse'
covid_case_num  = 'confirmed_7dav_incidence_num'
covid_case_prop = 'confirmed_7dav_incidence_prop'
start_day =  '2020-05-01'
end_day =    last_eval_date
colors = 1:4
```
### Forecasting results

We trained two forecasters, using forecast dates from October 1, 2021 to
(latest possible).  For each forecast date, we predict 5 days ahead to 
28 days ahead.  Predictions are made for the 23 CDC quantiles.
Training was done on the rates scale; one can transform the predictions to
the counts scale (as requested by the CDC) post hoc.

Based on the following results, we recommend using the "AR3+CHNG3" model, 
which uses three lags of the response and the Change flu indicator.  Adding
the lagged indicator provides modest improvements in WIS, and noticeable 
improvements in coverage (although all models are severely undercovering).

Recall that the Change Healthcare data being used here is "finalized", and
hence performance may be over-optimistic.  However, the pure autoregressive
model still exhibits noticeable improvement over the flat-line baseline model.

Finally, observe in the latter figures that especially at one-week ahead
(ahead=5), switching from the flat-line baseline to an autoregressive model
negatively impacts states with smaller population, while having neutral (mean WIS)
or positive (achieved 80% coverage) on states with larger population.  The
mechanics of how the forecasts change in states of varying population should
be more closely studied.

```{r plot_results_wis, fig.width=12, fig.height=12, warning=FALSE, message=FALSE}

full_results = readRDS('results/results_23.RDS')

results = full_results %>% filter (
      !stringr::str_ends(forecaster, '_7'),
      !stringr::str_detect(forecaster, 'logtrans'),
    ) %>% mutate (
      forecaster=stringr::str_replace(forecaster, '_23', ''),
    )

state_pop = covidcast::state_census %>% tibble (
    ) %>% transmute (
      geo_value = stringr::str_to_lower(ABBR),
      pop = POPESTIMATE2019,
    ) %>% mutate (
      state_with_pop = sprintf('%s, %d', geo_value, pop),
      state_with_pop = factor(state_with_pop,
                      levels = state_with_pop[order(pop)])
    )


results = results %>% left_join (
      state_pop,
      by='geo_value',
    ) %>% mutate (
      pop_wis = wis * pop,
      pop_ae = ae * pop,
    )

results$forecaster = factor(results$forecaster,
                            levels=c('Baseline','AR3','AR3+CHNG1','AR3+CHNG3'))

mean_results = results %>% group_by (
        forecaster,
        ahead,
      ) %>% summarize (
        mean_wis = mean(wis),
        mean_pop_wis = mean(pop_wis),
        mean_cov80 = mean(cov80),
        median_wis = median(wis),
        median_pop_wis = median(pop_wis),
      ) %>% ungroup
relative_mean = inner_join (
      mean_results %>% filter (
        forecaster == 'Baseline'
      ) %>% select (
        -forecaster,
      ),
      mean_results,
      by=c('ahead'),
      suffix=c('_Baseline', '')
    ) %>% transmute (
      ahead,
      forecaster,
      relative_mean_wis = mean_wis/mean_wis_Baseline,
      relative_median_wis = median_wis/median_wis_Baseline,
    )

plt_mean_wis = mean_results %>% ggplot (
      aes(x=ahead,
          y=mean_wis,
          color=forecaster)
    ) + geom_point (
    ) + geom_line (
    ) + scale_color_manual(
      values=colors,
    ) + geom_vline (
      xintercept=c(5, 12, 19, 26),
      linetype='dotted',
    ) + ylab (
      'Mean WIS'
    ) + theme(
      legend.position="bottom"
    )

plt_mean_pop_wis = mean_results %>% ggplot (
      aes(x=ahead,
          y=mean_pop_wis,
          color=forecaster)
    ) + geom_point (
    ) + geom_line (
    ) + scale_color_manual(
      values=colors,
    ) + geom_vline (
      xintercept=c(5, 12, 19, 26),
      linetype='dotted',
    ) + ylab (
      'Mean WIS (population-weighted)'
    ) + theme(
      legend.position="bottom"
    )

plt_relative_mean = relative_mean %>% ggplot (
      aes(x=ahead,
          y=relative_mean_wis,
          color=forecaster)
    ) + geom_point (
    ) + geom_line (
    ) + scale_color_manual(
      values=colors,
    ) + geom_vline (
      xintercept=c(5, 12, 19, 26),
      linetype='dotted',
    ) + ylab (
      'Mean WIS (relative to mean WIS of Baseline)'
    ) + theme(
      legend.position="bottom"
    )

plt_cov80 = mean_results %>% ggplot (
      aes(x=ahead,
          y=mean_cov80,
          color=forecaster)
    ) + geom_point (
    ) + geom_line (
    ) + scale_color_manual(
      values=colors,
    ) + geom_vline (
      xintercept=c(5, 12, 19, 26),
      linetype='dotted',
    ) + geom_hline (
      yintercept=0.8,
      linetype='dotted',
    ) + ylab (
      'Prop. 80% intervals that covered truth'
    ) + theme(
      legend.position="bottom"
    )

gridExtra::grid.arrange(plt_mean_wis,
                        plt_mean_pop_wis,
                        plt_relative_mean,
                        plt_cov80,
                        nrow=2)

```

```{r plot_joined_results_wis_diff, fig.width=12, fig.height=12, warning=FALSE, message=FALSE}
results_diff = results %>% filter (
      ahead %in% c(5, 12, 19, 26),
      forecaster %in% c('Baseline', 'AR3+CHNG3'),
    ) %>% group_by (
      geo_value,
      ahead,
      forecaster,
    ) %>% summarize (
      mean_wis = mean(wis),
      mean_cov80 = mean(cov80),
    ) %>% ungroup (
    ) %>% inner_join (
      state_pop,
      by='geo_value',
    )
joined_results_diff = inner_join (
      results_diff %>% filter (
        forecaster == 'Baseline',
      ) %>% select (
        -forecaster,
        -state_with_pop,
      ),
      results_diff %>% filter (
        forecaster == 'AR3+CHNG3',
      ) %>% select (
        -forecaster,
        -state_with_pop,
        -pop,
      ),
      by=c('geo_value', 'ahead'),
      suffix=c('_Baseline', '_AC3')
    ) %>% mutate (
      improved_mean_wis = mean_wis_AC3 <= mean_wis_Baseline,
      improved_mean_cov80 = mean_cov80_AC3 >= mean_cov80_Baseline,
    )

plt = ggplot(joined_results_diff,
             aes(x=pop,
                 y=mean_wis_Baseline,
                 xend=pop,
                 yend=mean_wis_AC3,
                 color=improved_mean_wis),
    ) + geom_segment(
      arrow=arrow(length=unit(0.1, 'cm')),
    ) + scale_color_manual (
      values = c(1, 4),
    ) + scale_color_manual(
      values=c('red', 'darkgreen'),
      name='AR3+CHNG3 improves over Baseline',
    ) + facet_wrap (
      vars(ahead),
      nrow=2
    ) + xlab (
      'Population'
    ) + ylab ( 
      'Mean WIS'
    ) + ggtitle (
      'Change in Mean WIS, from Baseline to AR3+CHNG3'
    ) + scale_x_log10 (
    ) + theme (
      legend.position='bottom',
    )

plt


plt = ggplot(joined_results_diff,
             aes(x=pop,
                 y=mean_cov80_Baseline,
                 xend=pop,
                 yend=mean_cov80_AC3,
                 color=improved_mean_cov80),
    ) + geom_segment(
      arrow=arrow(length=unit(0.1, 'cm')),
    ) + scale_color_manual (
      values = c(1, 4),
    ) + scale_color_manual(
      values=c('red', 'darkgreen'),
      name='AR3+CHNG3 improves over Baseline',
    ) + facet_wrap (
      vars(ahead),
      nrow=2
    ) + xlab (
      'Population'
    ) + ylab (
      'Prop. 80% intervals that covered truth'
    ) + ggtitle (
      'Change in 80% coverage, from Baseline to AR3+CHNG3'
    ) + scale_x_log10 (
    ) + theme (
      legend.position='bottom',
    )

plt
```
