# Epipredict Dogfood Yum 2023-07-05

We have match-up!

Previously:
- we were filtering the data before handing it to the forecasters manually
- on the right boundary, this doubled the effect of ahead on `train_end_date`
- on the left boundary, quantgen was using a training window, but epipredict was not; now we use n_training, and now we have sync

Now, we have:
- exact match-up in every state **AFTER** ensuring forecast values increase with quantile in the quantgen forecaster

To see this, run the code below. The variable `combined_out` contains the forecasts from both forecasters, the quantgen forecaster is `value` and epipredict is `value2`.

TODO: stress-test these findings, e.g.
- try different lags, aheads, and n_training
- try using an archive forecaster
- try a dataset where the data is not contiguous
- try a dataset where one geo has a different max time_value
- try a dataset where one geo has NAs
- try a dataset where one geo has a different min time_value

```{r}
library(epipredict)
library(epidatr)
library(tidyverse)
library(magrittr)
library(covidcast)
source(here::here("code", "quantgen.R"))
# Get data
data <- covidcast(
    "hhs",
    "confirmed_admissions_influenza_1d_prop_7dav",
    "state",
    "day",
    "*",
    epirange(20221201, 20230329),
) %>%
    fetch() %>%
    as.covidcast_signal(signal = "confirmed_admissions_influenza_1d_prop_7dav")
data %>% summarize(max(time_value), min(time_value))
data %>% tibble()
# Ensure same max time_value in every geo
data %>%
    group_by(geo_value) %>%
    summarize(max_time_value = max(time_value)) %>%
    filter(max_time_value != max(data$time_value))
# Ensure data contiguous in every geo
time_values <- seq.Date(as.Date("2022-12-01"), as.Date("2023-03-29"), by = "day")
data %>%
    group_by(geo_value) %>%
    summarize(contiguous = (setdiff(time_values, time_value) %>% length() == 0)) %>%
    filter(!contiguous)
# Ensure no NAs
data %>% summarize(sum(is.na(value)))
# Setup model args
n <- 4 * 7
lags <- c(0, 7, 14)
ahead <- 4
tau <- c(0.01, .025, 1:19 / 20, .975, .99)
forecast_date <- "2023-03-29"
states_dc_pr_vi <- c(
    "al", "ak", "az", "ar", "ca", "co", "ct", "dc", "de", "fl",
    "ga", "hi", "id", "il", "in", "ia", "ks", "ky", "la", "me",
    "md", "ma", "mi", "mn", "ms", "mo", "mt", "ne", "nv", "nh",
    "nj", "nm", "ny", "nc", "nd", "oh", "ok", "or", "pa", "ri",
    "sc", "sd", "tn", "tx", "ut", "vt", "va", "wa", "wv", "wi",
    "wy", "pr", "vi"
)
signals_ar <- tibble::tibble(
    data_source = "hhs",
    signal = "confirmed_admissions_influenza_1d_prop_7dav",
    start_day = 20200501,
    end_day = 20230329,
    geo_values = list(states_dc_pr_vi),
    geo_type = "state"
)
a <- ahead
dt <- lapply(list(lags), "-")
dt[[1]] <- c(dt[[1]], ahead)
df_list <- list(data)
df_wide <- covidcast::aggregate_signals(df_list, dt = dt, format = "wide")
train_end_date <- df_wide %>%
    select(time_value, tidyselect::starts_with(sprintf("value+%i:", a))) %>%
    tidyr::drop_na() %>%
    summarize(max(time_value)) %>%
    pull()
# Calculate predictions
out1 <- quantgen_forecaster(
    list(data),
    forecast_date,
    signals_ar,
    "day",
    ahead,
    "state",
    n,
    lags = lags,
    tau = tau,
    nonneg = TRUE
)
data2 <- data %>%
    select(geo_value, time_value, value) %>%
    as_epi_df()
out2 <- arx_forecaster(
    data2,
    "value",
    "value",
    quantile_reg(),
    args_list = arx_args_list(
        lags = lags,
        ahead = ahead,
        levels = tau,
        nonneg = TRUE,
        n_training = n
    )
)
# Compare predictions
ca_out1 <- out1[[1]] %>%
    tibble() %>%
    mutate(quantile = round(quantile, 5))
# Fix quantile ordering
ca_out1 <- ca_out1 %>%
    group_by(geo_value) %>%
    mutate(value = sort(value)) %>%
    ungroup()
ca_out2 <- out2$predictions %>%
    group_by(geo_value) %>%
    reframe(quantile = tau, value2 = quantile(.data$.pred_distn, tau)[[1]]) %>%
    mutate(quantile = round(quantile, 5))
combined_out <- ca_out1 %>%
    full_join(ca_out2, by = c("geo_value", "quantile"))
combined_out <- combined_out %>% mutate(
    abs_diff = abs(combined_out$value - combined_out$value2),
    rel_diff = abs(combined_out$value - combined_out$value2) / pmax(combined_out$value, combined_out$value2)
)
all.equal(combined_out$value, combined_out$value2)
```


# Epiprepdict Dogfood Yum 2023-04-26

We have identified two key data transformation steps that can serve as comparison checkpoints between
the production flu forecaster, i.e. `flu-hosp-forecast/code/quantgen.R`, and the epipredict AR forecaster,
i.e. `epipredict::arx_forecaster`:

1. lags and aheads calculation
2. training window filtering

In the production flu forecaster, we have

1. `covidcast::aggregate_signals` takes covidcast signal data and creates a "wide df" with key columns
   (geo_value, time_value) and value columns corresponding to each predictor's value combined with lags
   and aheads
    - e.g. a df with columns `geo_value` `time_value` `value+0:hhs_confirmed_admissions_covid_1d`
      `value-4:hhs_confirmed_admissions_covid_1d` `value+4:hhs_confirmed_admissions_covid_1d`
2. a filtering step that selects `n` most recent data points using
    - right bound: `max(time_value) - ahead`, where `max(time_value)` is taken across all geos
    - left bound: `max(time_value) - ahead - n + 1`, where `max(time_value)` is taken across all geos

The equivalent procedures in `epipredict::arx_forecaster` are

1. `step_epi_lag` and `step_epi_ahead`
    - the computation is done slightly differently, but after dropping NAs, the dataframes are the same
      in a few tested cases
2. `step_training_window` (though it's not built into arx_forecaster)
    - there appears to be a bug in `step_training_window`, where it doesn't group by geo when doing filtering

Results:

1. after omitting NAs, the outcomes of step (1) appear to be identical
2. after bugfixing `step_training_window`, we can do the next comparison step

```{r}
library(epipredict)
library(epidatr)
library(tidyverse)
library(magrittr)
library(covidcast)
library(recipes)
# Data setup
orig <- covidcast_signal(
    "hhs",
    "confirmed_admissions_covid_1d",
    geo_type = "state",
    geo_values = c("az", "al"),
    time_type = "day"
)
orig <- orig %>% tibble()
orig
small <- orig %>%
    filter(
        geo_value == "al" & (
            between(time_value, as.Date("2023-02-01"), as.Date("2023-02-07")) | between(time_value, as.Date("2023-02-14"), as.Date("2023-02-28"))
        ) |
            geo_value == "az" & between(time_value, as.Date("2023-02-01"), as.Date("2023-02-28"))
    )
small
small2 <- small %>%
    as_epi_df() %>%
    select(geo_value, time_value, value)
small2
# Gaps filled, but values before min(time_value) and after max(time_value) are not added
df_wide <- covidcast::aggregate_signals(list(small), list(c(0, -4, 4)), format = "wide") %>%
    as_tibble() %>%
    print(n = 20)
# Gaps filled and values before min(time_value) and after max(time_value) are added, however some between values not added
r <- epi_recipe(small2) %>%
    step_epi_lag(value, lag = c(0, 4)) %>%
    step_epi_ahead(value, ahead = 4)
df_wide2 <- r %>%
    prep(small2) %>%
    bake(new_data = NULL) %>%
    print(n = 20)
# The same test as above, but with both methods dropping NAs (which presumably the fitters will do themselves)
df_wide <- covidcast::aggregate_signals(list(small), list(c(0, -4, 4)), format = "wide") %>%
    as_tibble() %>%
    drop_na() %>%
    print(n = 20)
r <- epi_recipe(small2) %>%
    step_epi_lag(value, lag = c(0, 4)) %>%
    step_epi_ahead(value, ahead = 4) %>%
    step_naomit(all_predictors(), skip = FALSE) %>%
    step_naomit(all_outcomes(), skip = FALSE) %>%
    step_training_window(n_recent = 3)
df_wide2 <- r %>%
    prep(small2) %>%
    bake(new_data = NULL) %>%
    print(n = 27)
tib <- tibble::tibble(
    x = 1:10,
    y = 1:10,
    time_value = rep(seq(as.Date("2020-01-01"),
        by = 1,
        length.out = 5
    ), times = 2),
    geo_value = rep(c("ca", "hi"), each = 5)
) %>%
    as_epi_df()
epi_recipe(tib) %>%
    step_training_window(n_recent = 3) %>%
    recipes::prep(tib) %>%
    recipes::bake(new_data = NULL)
epi_recipe(tib) %>%
    step_naomit(skip = FALSE) %>%
    step_training_window(n_recent = 3) %>%
    recipes::prep(tib) %>%
    recipes::bake(new_data = NULL)
```
